Audiovisual Analytics Vocabulary and Ontology (AAVO): initial core and example expansion

### 2017-08-20 latex/article.tex
started article

%%-----------------------------------------------EDIT-----------------------------------------------
\title{NECESSARY ENHANCEMENTS OF LINKED DATA EXPRESSIVENESS FOR ONTOLOGIES}

%%-----------------------------------------------EDIT----------------------------------------------
\author
    {\rm \begin{tabular}{l} 
    \textbf{Renato Fabbri}$^{1}$ - {\textnormal renato.fabbri@gmail.com}\\%
%    \textbf{Marília M. Pisani}$^{2}$ - {\textnormal marilia.m.pisani@gmail.com}\\
    {\fontsize{11}{0}\selectfont $^{1}$University of São Paulo, Institute of Mathematical and Computer Sciences - São Carlos, SP, Brazil}\vspace*{-0.05cm} \\
%    {\fontsize{11}{0}\selectfont $^{2}$Federal University of ABC, Centre for Natural Sciences and Humanities - São Paulo, SP, Brazil}\vspace*{-0.05cm}\\
  \end{tabular}}
%%----------------------------------------------------------------------------------------------

@@

\begin{abstract}
	The semantic web has received many contributions of researchers as ontologies which,
	in this context, i.e. within RDF linked data, are formalized conceptualization that might use
	different protocols, such as RDFS, OWL DL and OWL FULL.
	In this article we describe new expressive techniques which were found necessary after
	elaborating dozens of OWL ontologies for the scientific academy, the State and the civil society.
	They consist in: 1) stating possible uses a property might have without incurring into axioms or restrictions;
	2) assigning a level of priority for an element (class, property, triple); 3) correct depiction in diagrams
	of relations between classes, between individuals which are imperative, and between individuals which are dispensable;
	4) association between OWL classes and SKOS concepts.
	We propose specific rules to accomplish these enhancements and exemplify both its use
	and the difficulties that arise because these techniques are currently not established as basic to the ontology designer.  
\end{abstract}

\keywords{\em{ Semantic web, Linked data, Ontology, RDFS, OWL}}

@@

\section{INTRODUCTION}
The semantic web is constituted by data which are linked in the same way
as web pages are: through HTTP and URLs.
W3C recommendations are the main sources of protocols and best practices of the field.
In practice, the terms 'linked data' and 'semantic web' are most often used interchangeably.
A distinction might arise in some contexts where on need to refer to the linked data or
the semantic web created by all or some portion of linked data, but, as a knowledge and technological
field, the terms are equivalent.

The semantic web is built using the Resource Description Framework (RDF).
The RDF data model is based on making statements in the form of triples
(``subject-predicate-object'') and using Unique Resource Identifier for
objects and concepts.
It is also part of the framework to use URIs that are URLs whenever possible,
to enable the data linkage.
Accordingly, one can write:
\begin{Verbatim}[fontsize=\footnotesize]
<http://example.org/people/mary>
	<http://example.org/properties/name> "Mary Shastacian" .
<http://example.org/people/mary>
	<http://example.org/properties/age> "57" .
<http://example.org/people/mary> 
	<http://example.org/properties/likes> 
                <http://example.org/concepts/reading> .
\end{Verbatim}
\noindent to express that there is person called Mary Shastacian which is 57 years old and likes reading.
There are many formats to write/serialize RDF data.
The example above is written in Turtle and we will use this format throughout this document.

One of the most important notions of the semantic web is that of an ontology.
An ontology, in this context, is a formalized conceptualizations, comprised by
concepts and relations between the concepts and between the relations themselves.
In current semantic web, most simple ontologies are written using the RDFS protocol,
by which one can specify, among other things:
\begin{itemize}
	\item concepts;
	\item properties, which are concepts that are used as predicates and thus relate concepts;
	\item special relations between concepts that state that one concept is mode general than the other\footnote{This
		kind of relation is called hypernymy. Examples: mammal is a hypernym of monkey, drink is a hypernym of beer.};
	\item the subjects and objects that can occurs in a triple where a specific property is a predicate.
\end{itemize}

One can also write an ontology by using OWL,
in which all the expressive capabilities of RDFS are
available, but one can also, among other things:
\begin{itemize}
	\item state ``property axioms'', i.e. specify if a property is e.g. reflexive or transitive;
	\item state ``class restrictions'', i.e. specify if a class instance e.g. necessarily hold a relation to another class instance or data.
\end{itemize}
OWL has a richer vocabulary than RDFS and is (way more) complex.
This complexity is a drawback together with the greater computational cost
for performing inference.
The advantage is the greater power to represent conceptualizations.

The semantic web practices evolves by constant practice and recommendations written
by many specialists usually after years of research and discussion.
Therefore, this document is written in a very humble manner,
without the pretense of yielding new recommendations abruptly,
but as a way to report issues and candidate workarounds/solutions
that emerged by years of intense practice in designing ontologies~\citep{pnud5,fabbri1,losd,ops}.

Section~\ref{sec:prob} is dedicated to exposing some of the problems
in current RDFS and OWL ontologies, with emphasis on the ones
to which this document holds proposals to resolve or lessen.
In Section~\ref{sec:prop} such proposals are described and discussed.
Final remarks are articulated in Section~\ref{sec:con}.

\section{THE PROBLEMS}\label{sec:prob}
The semantic web technology standards are very appreciated by the researchers, developers and users~\citep{semApr}.
Even so, there are many issues regarding current state of the art.
For example, if an individual of a class is found also to be a class,
it implies OWL Full which is a setting presently regarded as not computationally feasible.
This document is dedicated to proposals that resolve (or alleviate)
the problems described in the following subsections.

\subsection{No (standard) way to express potential relations}\label{sec:rel}
There is no recommendation about how to bind to a property a \emph{possible}
subject or object.
Using \texttt{rdfs:doman} and \texttt{rdfs:range} has the consequence that
\textbf{any} subject or object will have be a member of the class specified by
range or domain.
There are, however, many situations where this is not the case.
For example, one might find or create a property \texttt{example:name}.
Suppose it is intended to be used to express names of animals,
say because the person created the property while designing an ontology about animals.
If that is formalized in the property by \texttt{rdfs:domain},
it implies that whenever the \texttt{example:name} is used, the subject
will be inferred (e.g. by an automated reasoner) to be an animal.
Now suppose someone is describing a tool in RDF and wants to write its name.
If he/she uses the \texttt{example:name} property, it will be erroneously
expressing that the tool is an animal.
Such a person is required to create another property \texttt{example2:name}
to write the names.
What is worse: if the person wants to create an ontology about tools,
and wants to express that tools are likely, but not required, to have names,
the property \texttt{exampleX:name} with domain tool will have the same problem
as above, but now with tools instead of animals.
If \textbf{all} the tools have, without exception, a name, then one can express
it as an OWL existential class restriction.
But, again, that is often not the case. Example: algorithms are likely to have names,
but not all of them have names.
One workaround to this problem is to describe these potential relations as notes to the
classes and properties, but it does not favor the machine readable design of the semantic web.

\subsection{All classes and properties are equally fundamental}\label{sec:fun}
The relevance of this issue can be grasped by
tacking the expression of conceptualizations for humans.
Say someone wrote an ontology about machine learning and that it is very complete
in the sense that it contains many concepts, all those that the author could find.
It is straightforward to acknowledge that it will be very difficult for a newcomer
to make sense of everything.
Also, in many cases it is difficult for anyone to make sense of an ontology when it
comprises many concepts and relations.
Thus, it is very useful to have a standard way to express that some set of
concepts or triples are more fundamental then others in a layered fashion,
but there isn't a standard way.

\subsection{Distinction in diagrams of relations between classes and individuals and between imperative and optional relations}\label{sec:dia}
This is not a problem found in formalizing the ontology,
but a problem that arises when depicting the ontology as a diagram.
For example, the (class of) dinosaurs were succeeded by the (class of) mammals,
this is a relation between classes.
A specific animal might have siblings, this is a relation between individuals.
Also, a relation might be imperative for a class (i.e. it might be expressed as a class restriction)
or optional.
These are all poorly distinguished in diagrams that are meant to represent the ontologies.

\subsection{No obvious way to associate an OWL ontology and a SKOS vocabulary}
In a SKOS vocabulary, a concept is an individual of the class \texttt{skos:concept}.
In an OWL ontolgy, a concept is a class.
As stated in the beginning of this section, 
if an individual of a class is also a class,
one has an OWL Full context, which is
currently considered to require inference that is not computable.
This implies that one has to choose between making
an OWL ontology or a SKOS vocabulary or find a way to
make them compatible.
This might be the most complex issue tackled in this document.

\section{ENHANCEMENT PROPOSALS}\label{sec:prop}
As the problems stated in the previous section are distinct,
the proposed enhancements are described case-by-case.

\subsection{Potential domain, range, predicate and post predicate}
To address the problem described in Section~\ref{sec:rel},
we propose creating a new set of properties:
\begin{itemize}
	\item potential/nonrestrictive domain and range of a property might be expressed using\\
		\texttt{example:potentialDomain} and \texttt{example:potentialRange}.
	\item potential predicates when an instance of a class is a subject or an object might
		be expressed using\\
		\texttt{example:potentialPredicate} and \texttt{example:potentialPostPredicate}.
\end{itemize}

The \texttt{example:} in these triples should be replaced by a true namespace prefix, such
as \texttt{http://purl.org/socialparticipation/rdfe/}.
This was not done until the moment because we chose to wait a little longer,
present this article and distribute it online to find out if there is already some
solution currently in use and if there is some trustworthy maintainer that is willing
to keep these properties online in a more suitable namespace.

\subsection{Ascribing numeric values that reflect how fundamental is something}\label{sec:props}
To address the problem described in Section~\ref{sec:fun},
every property and class might me a subject to a triple:
\begin{Verbatim}[fontsize=\footnotesize]
	example:something example:fundamental 4 .
\end{Verbatim}
\noindent with ``4'' replaced by any other numerical value.
At first, it seems reasonable to assume that these values are
valid within the ontology and relative to each other.
Accordingly, one might ascribe \texttt{fundamental 100}
some kernel of meaning and lower values to other layers of
concepts and statements with more-or-less equal weight with respect
to being fundamental to the ontology.
As statements are usually not given a URI explicitly
in RDF files (one has to use ``RDF reification'' to give a triple a URI),
to bind a triple or a set of triples to a level,
one will probably be better of by using named graphs.
Reification is a solution, but a rather verbose and unusual one.

\subsection{Different arrow heads}
To address the problem described in Section~\ref{sec:dia},
we notice that: 1) the edges are often small, so using line styles
(such as dashed, dotted or full) is likely to yield bad results often.
The use of colored edges is an option, but given that many
journals require images to be black and white, it is not suitable
as the standard at least at the moment.
Now different arrow heads are in fact used (almost as a standard) 
as visual cues
to differentiate relations that are yield by properties from those
that associate a class to its superclass,
This is an indicative that using different arrow heads might
be a good strategy to reaching a good solution.
Also, the start of the edge might be a good visual cue to
discern the different type of relations.
We should recall here that in OWL a property might be specified to
satisfy a number of axioms, which care also prone to be represented
with the same artifices.

\subsection{OWL class as subclass of SKOS concept}\label{owlSkos}
There are different
strategies for making an OWL ontology and a SKOS vocabulary compatible~\citep{owlSkos}.
The solutions that are fit for general purposes, among the ones that were found or conceived by the author, are:
\begin{itemize}
	\item describe the ontology as OWL then translate it to SKOS, probably automated by means of a script.
	\item State that each OWL class is also a subclass of the \texttt{skos:Concept}.
		This yields a less redundant result, but it might not be in accordance to SKOS and OWL specifications
		or entail OWL Full.
		For example, the use of some SKOS properties might imply that the subject is an individual of the \texttt{skos:Concept} class.
		This solution needs to be further studied and discussed with specialized groups in order to be used with confidence.
\end{itemize}

The problem with the other solutions:
\begin{itemize}
	\item yields OWL Full (if a SKOS concept, an instance, is also an OWL class); or
	\item yields annotations on instances of OWL classes, not on the classes
		(if we annotate instances of OWL classes with SKOS properties); or
	\item yields a less formal ontology (if we derive the ontology from the vocabulary); or
	\item extrapolates the purpose of SKOS
		(giving enough expressive power to SKOS to achieve an OWL without degeneration of the conceptualization).
\end{itemize}

\section{CONCLUSIONS AND FURTHER WORK}\label{sec:con}
This article is a formalization of the issues to allow
the author and collaborators to confirm (or refute) that these problems
and the solutions herein proposed are sound and have not yet received
solutions by the semantic web community.
These issues and solutions are the result of years of dedication
to the subject, including learning by (online and in-person) classes
and designing dozens of OWL ontologies for academic and State purposes~\citep{pnud5}.

Next steps should include:
\begin{itemize}
	\item the pursue of feedback through circulation of this document (e.g. in online forums)
		to enhance the solutions herein proposed and better understand their possible uses (e.g. in reasoners).
	\item The use of these solutions in ontologies, both new ones an other that are already available.
	\item If these issues prove themselves relevant over time, we will consider contributing to the full realization of solutions.
		For example, we might help achieving and maintaining suitable URIs for the properties described in Section~\ref{sec:props}
		or help in the elaboration of convenient arrow heads for the problem described in Section~\ref{sec:dia}.
	\item The confirmation or refutation of the second possible solution to make OWL and SKOS compatible as described in Section~\ref{owlSkos}.
\end{itemize}

\subsection*{\textit{Acknowledgements}}
The author thanks CNPq for the funding received while researching the topic of this article,
the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
where we needed directions for investigation.

>>>>

\usepackage{grffile}

%%-----------------------------------------------EDIT-----------------------------------------------
\title{Audiovisual Analytics Vocabulary and Ontology (AAVO): initial core and example expansion}

%%-----------------------------------------------EDIT----------------------------------------------
\author
    {\rm \begin{tabular}{l} 
    \textbf{Renato Fabbri}$$ - {\textnormal renato.fabbri@gmail.com}\\%
    \textbf{Maria Cristina Ferreira de Oliveira}$$ - {\textnormal cristina@icmc.usp.br}\\
    {\fontsize{11}{0}\selectfont University of São Paulo, Institute of Mathematical and Computer Sciences - São Carlos, SP, Brazil}\vspace*{-0.05cm} \\
%    {\fontsize{11}{0}\selectfont $^{2}$Federal University of ABC, Centre for Natural Sciences and Humanities - São Paulo, SP, Brazil}\vspace*{-0.05cm}\\
  \end{tabular}}
%%----------------------------------------------------------------------------------------------

@@

\begin{abstract}
Visual Analytics might be defined as data mining though interactive visual interfaces.
The field and has received noticeable consideration by researchers and the industry. 
The literature, however, is complex because it involves many fields of knowledge
and is considerably recent.
In this article we describe an initial organization of this knowledge as an OWL ontology
and a SKOS vocabulary.
This effort might be useful in many ways that include conceptual considerations,
software implementations.
Within the results and discussions, we expose a core and an example expansion
of the conceptualization, and incorporate design issues that enhance the
expressive power of the abstraction. 
\end{abstract}

\keywords{\em{OWL, SKOS, Semantic web, Visual analytics, Data visualization}}

@@

\section{INTRODUCTION}
Visual Analytics' usual definition is data mining (or the science of analytical reasoning facilitated)
through interactive visual interfaces.
From this definition, one can grasp that at least three fields are directly related to visual analytics:
data mining, human-computer interaction, and data visualization.
Each one of these fields are multidisciplinary and known to be considerably complex
with multiple theories and vast literature.

This work is proposed as an organization of the knowledge related to visual analytics
as a SKOS vocabulary and an OWL ontology.
In other words, this document reports in an initial formalized conceptualization
in terms of linked data recommended technologies.
The uses this formalization can have includes:
\begin{itemize}
	\item introduction of the Visual Analytics subject for non-specialists.
	\item Expressing a concise overview.
	\item Queries.
	\item Relating objects (e.g. data and techniques) inside the field or between the field and other domains.
	\item Making inference about the concepts and objects to which they are related.
\end{itemize}

In the present case, where the formalization is done as linked data,
the conceptualization allows the queries, inferences, relations, etc.
to be performed also by machines.
Therefore, for example, a software might be able to relate a dataset
or analysis methods to visualization techniques to aid a user of
for automated reporting.

Section~\ref{sec:methods} uncover the methods.
Section~\ref{sec:res} is dedicated to stating and discussing the results.
Final remarks and future work are in Section~\ref{sec:con}.

\section{METHODS}\label{sec:methods}
The methods described here are standard of the semantic web
to achieve formalized conceptualizations.
The next sections addresses the subjects very briefly
and the interested reader can visit the literature
of the field and will surely find great items,
including the W3C recommendations~\citep{w3cld,ldb}.

\subsection{The semantic web}
The semantic web is constituted by data which are linked in the same way
as web pages are: through HTTP and URLs.
W3C recommendations are the main sources of protocols and best practices of the field.
In practice, the terms 'linked data' and 'semantic web' are most often used interchangeably.
A distinction might arise in some contexts where on need to refer to the linked data or
the semantic web created by all or some portion of linked data, but, as a knowledge and technological
field, the terms are equivalent.
The main topics of this article are data visualization (or visual analytics)
and semantic web.
Accordingly, all the sections tackle the subject of the semantic web.

\subsection{RDF}
The semantic web is built using the Resource Description Framework (RDF).
The RDF data model is based on making statements in the form of triples
(``subject-predicate-object'') and using Unique Resource Identifier for
objects and concepts.
It is also part of the framework to use URIs that are URLs whenever possible,
to enable the data linkage.
Accordingly, one can write:

\begin{Verbatim}[fontsize=\footnotesize]
<http://example.org/people/mary>
	<http://example.org/properties/name> "Mary Shastacian" .
<http://example.org/people/mary>
	<http://example.org/properties/age> "57" .
<http://example.org/people/mary> 
	<http://example.org/properties/likes> 
                <http://example.org/concepts/Reading> .
\end{Verbatim}
\noindent to express that there is person called Mary Shastacian which is 57 years old and likes reading.
There are many formats to write/serialize RDF data.
The example above is written in Turtle format and it will be used throughout this document.

In real settings, when everything is working as recommended,
each of these URIs (that are URLs!) might be accessed 
through HTTP to access more triples referring to the URI accessed.
From the triples above, one would be able to access triples
describing each of the properties: \texttt{example:properties/name}, \texttt{example:properties/age}, and\\
\texttt{example:properties/likes},
and the concept \texttt{example:concepts/Reading}.
In fact, the triples above would probably be available in the URI/URL:
\texttt{example:people/mary}.
One good working example of this is DBPedia~\citep{dbpedia}.
The process of accessing a URI to find more triples is called \emph{dereferencing the URI} (or simply dereferencing).

\subsection{RDFS}
The Resource Description Framework Schema (RDF Schema or simply RDFS)
is a set of classes and properties for the RDF data model that allows
basic descriptions of ontologies.
It supports taxonomic relations (hypernymy, i.e. relations stating that a concept is more general than another),
bindings of properties to objects and datatypes, and notes (label, comment, see also, is defined by).

\subsection{SKOS}
The Simple Knowledge Organization System (SKOS)
is a data model for representing controlled vocabularies.
SKOS is a W3C recommendation to facilitate publication
and use of vocabularies and is built upon RDF and RDFS.
It holds itself a vocabulary for concepts, notation, documentation,
semantic and mapping relations, and collections.

\subsection{OWL}
The Web Ontology Language (OWL) is a language for publishing ontologies on the web.
While RDFS holds basic relations necessary even for very rudimentary organization
knowledge and data, OWL is complex and allows one to formalize elaborate conceptualizations.
Using OWL, an ontology might have properties that are required to satisfy a number or axioms,
and classes that obey restrictions or e.g. are the result of the union of other classes.

\subsection{Interviews with specialists and literature consultation}
The standard approach to design an ontology, according to the literature,
is to interview specialists of the field to which the ontology is related,
or to absorb the established literature, or both.
This work is being developed using both approaches.
The second author is a data visualization specialist which was
interviewed by the first author.
Also, the first author is acquiring a deeper knowledge in the field.

\section{RESULTS AND DISCUSSION}\label{sec:res}
\subsection{AAVO core}
\begin{figure}[!htbp] %h or !htbp
\vspace{-2pt}
\begin{center}
% \includegraphics[height=6.7cm,width=9cm]{../figs/aavo0.01_minimum.png}
\includegraphics[width=\textwidth]{../figs/aavo0.01_minimum.png}
    \caption{}
\label{fig:minimum}%
\end{center}
\end{figure}

% potential enhancements, from the email to cristina

\subsection{AAVO example expansion}
\begin{figure}[!htbp] %h or !htbp
\vspace{-2pt}
\begin{center}
\includegraphics[width=\textwidth]{../figs/aavo0.01_exemplified.png}
    \caption{}
\label{fig:minimum}%
\end{center}
\end{figure}

\section{CONCLUSIONS AND FURTHER WORK}\label{sec:con}

\subsection*{\textit{Acknowledgements}}
The authors thank CNPq and FAPESP for the fundings received while researching the topic of this article,
the researchers of IFSC/USP and ICMC/USP for the recurrent collaboration in every situation
where we needed directions for investigation.
